Experience Report: Haskell in Computational Biology

### 2012-06-29 mrfy_experience_report.tex
put timestamp on paper
a pass over most of the paper.

@@

\newcommand\txprobj[3][]{a#1_{#2_{j-1}#3_j}}
\newcommand\txprobjj[3][]{a#1_{#2_{j-1}#3_j}}
\newcommand\alignwidth{\ensuremath C} % number of columns in an alignment
\newcommand\pairedwith[1]{{\pi(#1)}}

@@

@@

%\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{Haskell in Computational Biology}   % 'preprint' option specified.

@@

%
% It can't hurt to remind everyone involved exactly what an Experience
% Report is.
%
%
\begin{abstract}
%An Experience Report provides {evidence} that functional
%programming really works, or it describes obstacles that prevent
%functional programming from working.   
Haskell gives computational biologists
the flexibility and rapid prototyping of a scripting
language but performance comparable to~C++.
In~our experience, higher-order functions, lazy evaluation, and
monads really worked, but
profiling and debugging presented obstacles.
Haskell libraries vary greatly:
memoization combinators and parallel-evaluation
strategies helped us a lot,
but other, nameless libraries mostly got in our way.
Despite the obstacles and the uncertain quality of some libraries,
Haskell's ecosystem
made it easy for us to develop new algorithms in computational
biology.
\end{abstract}

@@

A~hidden Markov model carries probabilities on some states and on all
state transitions.
Both the probabilities and the states are determined by the alignment:
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}~$M_j$.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has amino acid~$x$ in
 column~$j$.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}~$I_j$.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained amino acid~$x$ by
insertion at column~$j$.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}~$D_j$.
The deletion state determines the probability that a homologous protein
has lost an amino acid  in column~$j$, i.e., that it has a gap.
\end{itemize}
The probabilities $e_{M_j}(x)$ and $e_{I_j}(x)$ are called emission probabilities.
The hidden Markov model also has distinguished \emph{Begin} and \emph{End} states.
In~our representation, each state contains a probability or a table of
probabilities, and it is also labelled with one of these labels:
\smallverbatiminput{statelabel}

@@

In a hidden Markov model, each transition is actually labeled with a
score, not a probability.
And we include the labels of the states that are
connected:
\smallverbatiminput{tprob}
In~a Plan7 hidden Markov model, as discussed above, 
there are seven transitions into each node:
\smallverbatiminput{tprobs}
In~our code, we select from this 7-tuple using the \texttt{aScore}
function;
\mbox{$\mathtt{aScore}\;s\;\hat s\;(j-1) = \txprobj s {{\hat s}}$},
where $s$~and~$\hat s$ form one of the 7~pairs above.

A~7-tuple of scores is included in the representation of each node in
the model.
This representation also includes tables of emission probabilities
$e_{M_j}$ and $e_{I_j}$.
These tables are read using function
\mbox{$\mathtt{eScore}\;s\;j\;i = e_{s_j}(x_i)$}.

@@

Another of us was curious to try coding Viterbi's equations
directly as recursive functions.
Like the classic recursive Fibonacci function, Viterbi's functions,
when implemented \naive ly,
perform extraordinarily badly.
But just like the Fibonacci function, Viterbi's functions can be
\emph{memoized}.
Our code implements a transformed version of Viterbi's equations, 
which operates on scores instead of on probabilities.
This version is
shown on the right-hand side of \figref{viterbi}.
The transformed equations minimize the
score (i.e., the negated log probability) for each combination of
node~$j$, amino 
acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.

Scores can be usefully paired with many types of values,
so we have defined a small abstraction:
\smallfuzzverbatiminput{10.8pt}{vscore}
Think of a value of type \texttt{Scored~a} as a container holding
 an~``\texttt a'' with a score written on the side.
The \texttt{/+/} function adds to the score without touching the
 container.

@@

The \texttt{Scored} abstraction makes it easy to implement
Viterbi's equations.
For example, to compute $V_j^{\prime M}(i)$ using the equation at the top
right of \figref{viterbi}, we call 
\mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}.
The equation adds~$e^{\prime}_{M_{j}}(x_{i})$, computed
with \texttt{eScore}, to a minimum.
The sum of an~$a'$~term and a $V'_{j-1}(i-1)$ term 
is computed by function \texttt{avSum},
the terms are computed by \texttt{aScore} and \texttt{vee''}, respectively.
\smallverbatiminput{vfix}
The right-hand side is wrapped in a call to
\mono{\mbox{fmap (Mat `cons`)}}.
This call performs a computation \emph{not} shown in \figref{viterbi}:
\mrfy\ not only computes the probability of the most likely path but 
also computes the path itself.
Function \texttt{cons} adds~$M$ to a path;
we~avoid primitive~\primcons\ for reasons explained
in \secref{performance} below.

@@

Memoization makes \verb+vee'+ perform just as well as the classic
dynamic-programming code.
And~the call to \texttt{Memo.memo3} is the \emph{only} part of the code
devoted to dynamic programming.
By~contrast, standard implementations of the Viterbi algorithm, such as in HMMER,
spend much of their code 
managing dynamic-programming tables.
Haskell enabled us write simple, performant code with little effort.
%
Because the memoized version so faithfully resembles the equations in
\figref{viterbi}, we~retired the classic dynamic-programming version.

@@

Our Haskell code for hidden Markov models and Viterbi's algorithm
solves the same problems as existing C++~code.
Other researchers using Haskell may also have to reimplement code,
but researchers may also reimplement code out of choice, not
necessity.
For~example, both SMURF and HMMER contain new implementations of these
same algorithms.
In~MRFy, we are using the Haskell implementations to help solve new
problems.

@@

The new terms in Viterbi's equations are accompanied by a new model.
Within a beta strand, amino acids are not inserted or deleted, so
a bonded pair of beta strands is modeled by
 a pair of sequences of match states.
Between the beta strands, the model is structured as before.
The~combined model, an~example of which is shown in \figref{mrf}, is called a
\textit{Markov random field}.

MRFy treats the beta 
strands in the model
 as ``beads'' which can be slid along the query sequence.
A~positioning of the beta strands is called
 a \emph{placement}.
A~placement's probability is computed
based on observed frequencies of paired amino acids in 
hydrogen-bonded beta strands~\cite{Cowen:2002p588}.
Given a placement, the maximum likelihood of the remaining parts of
the query sequence, which appear between and around beta strands,
can be computed quickly and exactly 
using Viterbi's algorithm.
The exact result is a probability that is
\emph{conditioned} on the placement.

MRFy searches for good placements stochastically.
\mrfy\ implements
random hill
climbing, simulated  
annealing, multistart simulated annealing, and a genetic algorithm.
These algorithms share much code, and they use higher-order functions,
existentially 
quantified types, and lazy evaluation.

We~describe \mrfy's search abstractly: \mrfy\ computes a sequence
of \emph{points} in a search space.
A~point is typically a single placement or perhaps a collection of
placements. 
Each point also has a \texttt{Score}; \mrfy\ looks for points with
good scores.

Ideally, MRFy would use now-classic lazy, modular
technique advocated by \citet{hughes:why}, in which one function
computes an infinite sequence of points, and another function
uses a finite prefix to decide on an approximation.
But because MRFy's search is stochastic, 
making MRFy's search lazy and modular is not so easy.

To illustrate the difficulties, we~discuss our simplest search:
random hill climbing.
%The space of beta-strand placements has no notion of slope or
%gradient,
From any given point in the search space, this search moves a random
distance in a random 
direction.
If~the move leads to a better point, it
is \emph{useful};
otherwise it is \emph{useless}.
\smallverbatiminput{utility}
(We~also use \texttt{Useful} and \texttt{Useless} to tag \emph{points}.)
With luck, an~infinite sequence of useful moves converges
at a local optimum.

@@

We~address this problem by combining ``generate and filter'' into a
single abstraction, which has type \mono{SearchGen pt r}.
Type variable~\texttt{pt} is a point in the search
space, and \texttt{r} is a
random-number generator.
\mono{Rand~r} is a monad of stochastic computations:
\smallverbatiminput{gen.tex}
The monadic computation \texttt{pt0} randomly selects a starting point
for search;
\texttt{nextPt} produces a new point from
an existing point.
Because scoring can be expensive, both \texttt{pt0}
and \texttt{nextPt} use \emph{scored} points, and they can
reuse scores from previous points.

To tell if a point returned by \texttt{nextPt} is useful, we call
 the \texttt{utility} function,
which scrutinizes a move represented as follows:
\smallverbatiminput{move}
The decision about utility uses not only a source of randomness but
also the \emph{cumulative cost} of the point, which we define to be
the number of points explored previously.
The~cumulative cost of the current point is also the age of the
search,
and
in~simulated annealing, for example, as the search ages,
the \texttt{utility} function becomes less likely to accept a move
that results in a worse score.

\noindent
\vbox{
Using these pieces, function \texttt{everyGen} produces an infinite
sequence containing a mix of useful and useless moves:
\smallfuzzverbatiminput{10.8pt}{everygen}
}
\noindent
Because both \texttt{nextPt} and \texttt{utility} consume randomness,
the computation is very sequential, but it does use a little bit of
laziness:
from its starting point, \texttt{everyPt} provides an infinite list of
randomly chosen successor points, then calls \texttt{costedUtility} to
tag each one with the search's age and the point's utility.
And the computation of \texttt{successors} should explain why we
separate \texttt{pt0}~from~\texttt{nextPt} instead of using a single
function that produces an infinite list:
We~don't \emph{want} the infinite list that would result from
applying \texttt{nextPt} many times in succession;
we~want the infinite lists that results from applying \texttt{nextPt}
to \texttt{startPt} many times in succession, each time with a
different source of randomness.

Once the successors have been computed and tagged,
\texttt{span} finds the first useful successor.
In~case there \emph{is} no successor, \texttt{everyPt} also returns
all the useless successors.
\remark{Story about the killer strict Rand monad? Where?}
If~we do find a useful successor, we start searching anew from that
point, with a recursive call to \texttt{everyPt}.
(Because \texttt{everyPt} is monadic, the points accumulated so far are
appended to its result using the \verb+<$>+~operator.)
%
The most informative part of \texttt{everyPt} is last expression of
the \texttt{do} block,
which shows that the result begins with a useful point,
is followed by a (possibly infinite, possibly empty) list of useless
points, and then continues recursively with the results of \texttt{everyPt}.

@@

The definition of \texttt{SearchStop} reveals the forms of
non-modularity which are inherent in \mrfy's search algorithm.
First, we need \texttt{Utility} because if we omit the useless states,
search might not terminate.
Second, we need \texttt{CCosted} because some of our test functions
decide to terminate based either on the cumulative cost of the most
recent point or on the difference between costs of successive points.

@@

At each point in its search,
MRFy calls \verb+vee'+ several
 times.
Our~\verb+vee'+
function computes a \mbox{\small \texttt{Scored [StateLabel]}}, that~is, 
an~optimal path and its score.
But at intermediate points in \mrfy's search,
\mrfy\ uses only the
score.
Even though Haskell evaluation is lazy, \verb+vee'+ 
still allocates thunks that could compute paths.
To~measure the relevant overhead,
we~cloned \verb+vee'+ and modified it to compute
only a score, with no path.
This change improved 
run time by nearly~50\%.
%%But when \texttt{search} finally decides on a placement, we need a
%%version of \verb+vee'+ that \emph{does} compute a path, so we can
%%extract the optimal parse for that placement.

@@

We~found another opportunity to improve performance through parallelism.
Calls on query sequences between beta strands are independent,
as are calls on different placements.
We~found them easy to parallelize
using
\texttt{Control.Parallel}:
we substituted
\texttt{parmap rseq} for \texttt{map}.\remark{Did NR break this?}
Even when there is just one placement per point,
MRFy can keep 6~cores busy.
\remark{This paragraph needs to talk about \textbf{speedups}.}
Given multiple placements per point,
it~can use 48~cores.
However, we have found that MRFy performs best, at least on our
hardware, with between 12 and 16 cores.
We have access to only
one 48-core system, and while MRFy will keep all 48 cores busy, it
performs only slightly better, at an average of 35 seconds to find an alignment
on a large\remark{QUANTIFY!!} protein structure, than when given 16
cores, which takes an average 
of 37 seconds.

@@

We hate the code we've written to help debug
run-time errors.
Our~group's legacy file
format is poorly documented: when beta
strands overlap and are doubly paired, the invariants of the
format are not clear.
This format is hard to deal with no matter what
programming language is used, and faults are to be expected.
But using Haskell, we had a hard time diagnosing run-time errors.
Calls to
\texttt{trace} 
littered our code,
even when relegated to wrapper functions.
We~weren't aware of the backtrace feature of GHC's profiler,
and even after we discovered that feature, it didn't help:
the profiler can be used only on very small test cases that didn't always trigger
run-time errors.
This same limitation affected the GHCi debugger; 
in~GHCi, our \texttt{vee'} function
is too slow to be runnable on nontrivial
inputs.
Moreover, the GHCi debugger can only set breakpoints at top-level functions
or at specific column/line positions, which made debugging the memoized
\texttt{vee''} function impractical.
In~August 2011, 
Lennart Augustsson said that the biggest advantage of Strict
Haskell is getting a stack trace on error,
and Simon Marlow said that he may have figured out how to track call
stacks properly in a lazy functional language.
We~can't wait.

Our difficulties with debugging led to significant disagreements on
our team.
The two junior members wanted to apply the debugging
skills they had honed through years of imperative programming.
These skills did not transfer well to Haskell.
The senior member of the team kept saying that any proper approach to
debugging Haskell code should involve QuickCheck.
This advice was not helpful.

@@

Our experience was that only the senior member of our team was able to use
QuickCheck easily.
In~retrospect, we~have identified some obstacles to using
QuickCheck.
\begin{itemize}
\item
The examples and tutorials we found focused 
predominantly on writing and testing properties using data types that
already implemented class \texttt{Arbitrary}.
We~didn't understand the \texttt{Arbitrary} class very well, 
perhaps because the overloaded \texttt{arbitrary} value is not a
function.
For programmers who have seen lot of object-oriented dispatch
on \emph{arguments}, it is hard to grasp how the appropriate version
of \texttt{arbitrary} is chosen by Haskell's type-class system.
\item
Our difficulties were compounded by a weak understanding of monads.
We~were too baffled by QuickCheck's \texttt{Gen} monad to grasp its
true importance.
\item
We~continually overlooked the critical role of \emph{shrinking}.
As~a result, on the one or two occasions we did uncover a problem
using QuickCheck, we got witnesses that were much too large to be
informative. 
\end{itemize}
Because of these obstacles, 
we wrote thousands of lines of code without ever defining an instance
of \texttt{Arbitrary} and therefore
without looking hard at QuickCheck.
After the fact, we were overwhelmed by the work
involved in writing and testing instances
of
\texttt{Arbitrary}.
The work got done only because the whole team got involved in order to
meet the deadlines for this paper.

@@

When performance has mattered, members of our group, like other
computational biologists, have used~C++.
To~compare our Haskell experience with our C++ experience,
we discuss three tools:
\begin{itemize}
\item
Matt~\cite{Menke:2008wu} is used to create alignments like that shown
in \figref{alignment}.
It~comprises about 12,000 lines of~C++.
The only external tools or libraries it uses are \texttt{zlib}
and \texttt{OpenMP}, and its initial development took two years.
\item
SMURF
\cite{Menke:2010ti} is used to detect homologous proteins in the presence
of paired beta strands.
It~comprises about 9,000 lines of~C++, of which about 1,000~lines are
shared with~Matt.
It~uses no external tools or libraries, and
its initial development took a year and a half.
It uses multidimensional dynamic programming to exactly compute the alignments
for which MRFy relies on stochastic search.
As a result, in the presence of complex beta-strand topologies, SMURF is
computationally intractable.
\item
MRFy is used to detect homologous proteins in the presence of paired
beta strands; it effectively supplants SMURF.
It~comprises about 1,200 lines of~Haskell.
It~uses several external tools and libraries, of which the most
notable are Parsec, the BioHaskell bioinformatics library, and
the libraries \path{Data.MemoCombinators}, \path{Control.Parallel},
and \path{Data.Vector}.
Its initial development took about three months.
%%%% \remark{Others were mostly full-time work, but no need to distinguish.}
\end{itemize}
Like much research software, all three of these tools were written
 in~haste.
We~have experience modifying the older tools.

@@

Ideally we would use QuickCheck to find bugs, but as we mention in
\secref{awkward-quickcheck}, we found obstacles.
We~have now overcome these obstacles, but
we~sorely regret not doing so earlier.
In~light of our experience, we have instituted a new programming practice:
whenever we introduce a new data type,
we~write the instance of
\texttt{Arbitrary} right away, while
relevant invariants are still fresh in memory.
When invariants are not enforced by Haskell's static type system,
we write them as Haskell predicates.
We can then immediately run QuickCheck on each predicate, to verify
that our Arbitrary instance agrees with the predicate.
For each predicate~\texttt{p} we can also check
\mono{fmap (all p . shrink) arbitrary}.

@@

You~also need to know that while the Haskell community offers many
enticing tools, libraries, and packages,
not all of them are worth using.
Some are not ready for prime time, and some were once great but are
no~longer maintained.
The great packages, like \texttt{Data.MemoCombinators} and Parallel
Strategies, are truly great.
But~for amateurs, it's not always easy to tell the great packages from the wannabes
and the has-beens.
And even some of the great packages could be better documented, with
more examples.

>>>>

\newif\iftimestamp\timestamptrue  % show MD5 stamp of paper

% \timestampfalse % it's post-submission time

\IfFileExists{timestamp.tex}{}{\timestampfalse}

@@

\newcommand\txprobj[3][]{a#1_{{#2}_{j-1}{#3}_j}}
\newcommand\txprobjj[3][]{a#1_{{#2}_{j-1}{#3}_j}}
\newcommand\alignwidth{\ensuremath C} % number of columns in an alignment
\newcommand\pairedwith[1]{{\pi(#1)}}

@@

\IfFileExists{timestamp.tex}{\input{timestamp}}{}
\renewcommand{\floatpagefraction}{0.9} % must be less than \topfraction
\renewcommand{\topfraction}{0.95}
\renewcommand{\textfraction}{0.05}

@@

%\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{Haskell in Computational Biology}   % 'preprint' option specified.
\iftimestamp
\preprintfooter{\mdfivestamp} 
\fi

@@

%
% It can't hurt to remind everyone involved exactly what an Experience
% Report is.
%
%
\begin{abstract}
%An Experience Report provides {evidence} that functional
%programming really works, or it describes obstacles that prevent
%functional programming from working.   
Haskell gives computational biologists
the flexibility and rapid prototyping of a scripting
language but performance comparable to~C++.
In~our experience, higher-order functions, lazy evaluation, and
monads really worked, but
profiling and debugging presented obstacles.
Also, Haskell libraries vary greatly:
memoization combinators and parallel-evaluation
strategies helped us a lot,
but other, nameless libraries mostly got in our way.
Despite the obstacles and the uncertain quality of some libraries,
Haskell's ecosystem
made it easy for us to develop new algorithms in computational
biology.
\end{abstract}

@@

A~hidden Markov model carries probabilities on some states and on all
state transitions.
Both the probabilities and the states are determined by the alignment:
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}~$M_j$.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has amino acid~$x$ in
 column~$j$.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}~$I_j$.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained amino acid~$x$ by
insertion at column~$j$.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}~$D_j$.
The deletion state determines the probability that a homologous protein
has lost an amino acid by deletion from column~$j$.
\end{itemize}
The probabilities $e_{M_j}(x)$ and $e_{I_j}(x)$ are \emph{emission probabilities}.

A hidden Markov model also has distinguished \emph{Begin} and \emph{End} states.
In~our representation, each state contains a probability or a table of
probabilities, and it is also labeled with one of these labels:
\smallverbatiminput{statelabel}

@@

In a hidden Markov model, each transition is actually labeled with a
score, not a probability.
And in~a Plan7 model, as discussed above, 
there are seven transitions into each node:
\smallverbatiminput{tprob-tprobs}
%\smallverbatiminput{tprobs}
In~our code, we select from this 7-tuple using \texttt{aScore}, a
function which satisfies
\mbox{$\mathtt{aScore}\;s\;\hat s\;(j-1) = \txprobj s {\hat s}$},
where $s$~and~$\hat s$ form one of the 7~pairs above.
Such a 7-tuple of scores is stored in each node in
the model.
A~node also contains tables of emission probabilities
$e_{M_j}$~and~$e_{I_j}$.
These tables are read using function
\mbox{$\mathtt{eScore}\;s\;j\;i = e_{s_j}(x_i)$}.

@@

Another of us was curious to try coding Viterbi's equations
directly as recursive functions.
Like a recursive Fibonacci function, Viterbi's functions,
when implemented \naive ly,
take exponential time.
But like the Fibonacci function, Viterbi's functions can be
\emph{memoized}.
Our code implements a transformed version of Viterbi's equations, 
which operates on scores, not on probabilities.
This version is
shown on the right-hand side of \figref{viterbi}.
The transformed equations minimize the
score (the negated log probability) for each combination of
node~$j$, amino 
acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.

Scores can be usefully paired with many types of values,
so we have defined a small abstraction:
\smallfuzzverbatiminput{10.8pt}{vscore}
Think of a value of type \mono{Scored~a} as a container holding
 an~``\texttt a'' with a score written on the side.
The \texttt{/+/} function adds to the score without touching the
 container.

@@

The \texttt{Scored} abstraction makes it easy to implement
Viterbi's equations.
For example, to compute $V_j^{\prime M}(i)$ using the equation at the top
right of \figref{viterbi}, we call 
\mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}.
The equation adds~$e^{\prime}_{M_{j}}(x_{i})$, computed
with \texttt{eScore}, to a minimum of sums.
The sum of an~$a'_{s\hat s}$~term and a $V^{\prime s}_{j-1}(i-1)$ term 
is computed by function \texttt{avSum}, in which
the terms are computed by \texttt{aScore} and \texttt{vee''}, respectively:
\smallverbatiminput{vfix}
%%%%%%%%%% horrible line break
% The right-hand side of \verb+vee'+ is wrapped in a call to
What about the call to
\mono{\mbox{fmap (Mat `cons`)}}?
This call performs a computation \emph{not} shown in \figref{viterbi}:
\mrfy\ computes not only the {probability} of the most likely path but 
also the actual path itself.
Function \mono{(Mat `cons`)} adds~$M$ to a path;
we~avoid \mbox{\mono{(Mat :)}} for reasons explained
in \secref{performance} below.

@@

Memoization makes \verb+vee'+ perform as well as our classic
dynamic-programming code.
And~the call to \texttt{Memo.memo3} is the \emph{only} part of the code
devoted to dynamic programming.
By~contrast, standard implementations of Viterbi's algorithm, such as in HMMER,
spend much of their code 
managing dynamic-programming tables.
Haskell enabled us write simple, performant code with little effort.
%
Because the memoized version so faithfully resembles the equations in
\figref{viterbi}, we~retired the classic version.

@@

Our Haskell code for hidden Markov models and Viterbi's algorithm
solves the same problems as existing C++~code.
\remark{Drop this \P\ or move it}
Other researchers using Haskell may also have to reimplement code,
but researchers may also reimplement code out of choice, not
necessity.
IN COMPUTATIONAL BIOLOGY, REIMPLEMENTING EXISTING ALGORITHMS IS UNREMARKABLE.
For~example, both SMURF and HMMER contain new implementations of these
same algorithms.
In~MRFy, we are using the Haskell implementations to help solve new
problems.

@@

The new terms in Viterbi's equations are accompanied by a new model.
Within a beta strand, amino acids are not inserted or deleted, so
a bonded pair of beta strands is modeled by
 a pair of sequences of match states.
Between beta strands, the model is structured as before.
The~combined model, an~example of which is shown in \figref{mrf}, is called a
\textit{Markov random field}.

MRFy treats the beta 
strands in the model
 as ``beads'' which can slide along the query sequence.
A~positioning of the beta strands is called
 a \emph{placement}.
A~placement's probability is computed
based on observed frequencies of paired amino acids in 
hydrogen-bonded beta strands~\cite{Cowen:2002p588}.
Given a placement, the maximum likelihood of the remaining parts of
the query sequence, which appear between and around beta strands,
can be computed quickly and exactly 
using Viterbi's algorithm.
The exact result is a probability that is
\emph{conditioned} on the placement.

MRFy searches for good placements stochastically.
\mrfy\ implements
random hill
climbing, simulated  
annealing, multistart simulated annealing, and a genetic algorithm.
These algorithms share much code, and \mrfy\ implements them using
higher-order functions, 
existentially 
quantified types, and lazy evaluation.

We~describe \mrfy's search abstractly: \mrfy\ computes a sequence
of \emph{points} in a search space.
The type of point is existentially quantified, but it is typically
a single placement or perhaps a population of placements. 
Each point also has a \texttt{Score}; \mrfy\ looks for points with
good scores.

Ideally, MRFy would use now-classic, lazy, modular
technique advocated by \citet{hughes:why}, in which one function
computes an infinite sequence of points, and another function
uses a finite prefix to decide on an approximation.
But because MRFy's search is stochastic, 
making MRFy's search lazy and modular is not so easy.

To illustrate the difficulties, we~discuss our simplest search:
random hill climbing.
%The space of beta-strand placements has no notion of slope or
%gradient,
From any given point in the search space, this search moves a random
distance in a random 
direction.
If~the move leads to a better point, 
we call it \emph{useful};
otherwise it is \emph{useless}.
\smallverbatiminput{utility}
(We~also use \texttt{Useful} and \texttt{Useless} to tag \emph{points}.)
With luck, an~infinite sequence of useful moves converges
at a local optimum.

@@

We~address this problem by combining ``generate and filter'' into a
single abstraction, which has type \mono{SearchGen pt r}.
Type variable~\texttt{pt} is a point in the search
space, and \texttt{r} is a
random-number generator.
\mono{Rand~r} is a lazy monad of stochastic computations:
\smallverbatiminput{gen.tex}
The monadic computation \texttt{pt0} randomly selects a starting point
for search;
\texttt{nextPt} produces a new point from
an existing point.
Because scoring can be expensive, both \texttt{pt0}
and \texttt{nextPt} use \emph{scored} points, and they can
reuse scores from previous points.

To tell if a point returned by \texttt{nextPt} is useful, we call
 the \texttt{utility} function,
which scrutinizes a move represented as follows:
\smallverbatiminput{move}
The decision about utility uses not only a source of randomness but
also the \emph{cumulative cost} of the point, which we define to be
the number of points explored previously.
The~cumulative cost of the current point is also the age of the
search,
and
in~simulated annealing, for example, as the search ages,
the \texttt{utility} function becomes less likely to accept a move
that worsens the score.

\noindent
\vbox{
Using these pieces, function \texttt{everyPt} produces an infinite
sequence containing a mix of useful and useless moves:
\smallfuzzverbatiminput{10.8pt}{everygen}
}
\noindent
Both \texttt{nextPt} and \texttt{utility} are monadic, but 
we can still use a little bit of
laziness:
from its starting point, \texttt{everyPt} produces an infinite list of
randomly chosen successor points, then calls \texttt{costedUtility} to
tag each one with a cumulative cost and a utility.
If~you look carefully at how \texttt{successors} is computed,
we~hope you will understand why we
separate \texttt{pt0}~from~\texttt{nextPt} instead of using a single
function that produces an infinite list:
We~don't \emph{want} the infinite list that would result from
applying \texttt{nextPt} to many points in succession;
we~want the infinite list that results from applying \texttt{nextPt}
to \texttt{startPt} many times in succession, each time with a
different source of randomness.

Once the successors have been computed and tagged,
\texttt{span} finds the first useful successor.
In~case there \emph{is} no successor, \texttt{everyPt} also returns
all the useless successors.
If~we do find a useful successor, we start searching anew from that
point, with a recursive call to \texttt{everyPt}.
(Because \texttt{everyPt} is monadic, the points accumulated so far are
appended to its result using the \verb+<$>+~operator.)
%
The most informative part of \texttt{everyPt} is last expression of
the \texttt{do} block,
which shows that the result begins with a useful point,
is followed by a (possibly infinite, possibly empty) list of useless
points, and then continues recursively with the results of \texttt{everyPt}.

@@

The definition of \texttt{SearchStop} reveals two forms of
non-modularity which are inherent in \mrfy's search algorithm.
First, we need \texttt{Utility} because if we omit the useless states,
search might not terminate.
Second, we need \texttt{CCosted} because some of our test functions
decide to terminate based either on the cumulative cost of the most
recent point or on the difference between costs of successive points.

@@

At each point in its search,
MRFy calls \verb+vee'+ several
 times.
Our~\verb+vee'+
function computes a \mbox{\mono{Scored [StateLabel]}}, that~is, 
an~optimal path and its score.
But at intermediate points in \mrfy's search,
\mrfy\ uses only the
score.
Even though Haskell evaluation is lazy, \verb+vee'+ 
still allocates thunks that could compute paths.
To~measure the relevant overhead,
we~cloned \verb+vee'+ and modified it to compute
only a score, with no path.
This change improved 
run time by nearly~50\%.
%%But when \texttt{search} finally decides on a placement, we need a
%%version of \verb+vee'+ that \emph{does} compute a path, so we can
%%extract the optimal parse for that placement.

@@

We~also found an opportunity to improve performance through parallelism.
Calls on query sequences between beta strands are independent,
as are calls on different placements.
We~found them easy to parallelize
using
\texttt{Control.Parallel}:
we substituted
\texttt{parmap rseq} for \texttt{map}.
Even when there is just one placement per point,
MRFy can keep 6~cores busy.
\remark{This paragraph needs to talk about \textbf{speedups}.}
Given multiple placements per point,
it~can use 48~cores.
However, we have found that MRFy performs best, at least on our
hardware, with between 12 and 16 cores.
We have access to only
one 48-core system, and while MRFy will keep all 48 cores busy, it
performs only slightly better, at an average of 35 seconds to find an alignment
on a large\remark{QUANTIFY!!} protein structure, than when given 16
cores, which takes an average 
of 37 seconds.

@@

We hate the code we've written to help debug
run-time errors.
Our~group's legacy file
format is poorly documented: when beta
strands overlap and are doubly paired, the invariants of the
format are not clear.
This format is hard to deal with no matter what
programming language is used, and faults are to be expected.
But using Haskell, we had a hard time diagnosing run-time errors.
Calls to
\texttt{trace} 
littered our code,
even when relegated to wrapper functions.
We~weren't aware of the backtrace feature of GHC's profiler,
and even after we discovered that feature, it didn't help:
the profiler can be used only on very small test cases that didn't always trigger
run-time errors.
This same limitation affected the GHCi debugger; 
in~GHCi, our \texttt{vee'} function
is too slow to be runnable on nontrivial
inputs.
Moreover, the GHCi debugger can set breakpoints only at top-level functions
or at specific column/line positions, which made debugging the memoized
\texttt{vee''} function impractical.
In~August 2011, 
Lennart Augustsson said that the biggest advantage of Strict
Haskell is getting a stack trace on error,
and Simon Marlow said that he may have figured out how to track call
stacks properly in a lazy functional language.
We~can't wait.

Our difficulties with debugging led to internal disagreements.
The~junior members of our team wanted to apply the debugging
skills they had honed through years of imperative programming.
These skills did not transfer well to Haskell.
The senior member of the team kept saying that any proper approach to
debugging Haskell code should involve QuickCheck.
This exhortation was unhelpful.

@@

Only the senior member of our team was able to use
QuickCheck easily.
In~retrospect, we~have identified some obstacles that prevented the
junior people from using
QuickCheck.
\begin{itemize}
\item
The examples and tutorials we found focused 
predominantly on writing and testing properties using data types that
already implemented class \texttt{Arbitrary}.
We~didn't understand the \texttt{Arbitrary} class very well, 
perhaps because the overloaded \texttt{arbitrary} value is not a
function.
For programmers accustomed to object-oriented dispatch
on \emph{arguments}, it is hard to grasp how Haskell's type-class
system can find the proper version
of \texttt{arbitrary}.
\item
Our difficulties were compounded by a weak understanding of monads.
We~were too baffled by QuickCheck's \texttt{Gen} monad to grasp its
importance.
\item
We~continually overlooked the critical role of \emph{shrinking}.
As~a result, on the one or two occasions we did 
use QuickCheck, the counterexamples were too large to be
informative. 
\end{itemize}
Because of these obstacles, 
we wrote thousands of lines of code without ever defining an instance
of \texttt{Arbitrary} and therefore
without looking hard at QuickCheck.
After the fact, we were overwhelmed by the work
involved in writing and testing instances
of
\texttt{Arbitrary}.
The work got done only because the whole team got involved in order to
meet the deadlines for this paper.

@@

When performance has mattered, members of our group, like other
computational biologists, have used~C++.
To~compare our Haskell experience with our C++ experience,
we discuss three tools:
\begin{itemize}
\item
Matt~\cite{Menke:2008wu} is used to create alignments like that shown
in \figref{alignment}.
It~comprises about 12,000 lines of~C++.
The only external tools or libraries it uses are \texttt{zlib}
and \texttt{OpenMP}, and its initial development took two years.
\item
SMURF
\cite{Menke:2010ti} is used to detect homologous proteins in the presence
of paired beta strands.
It~comprises about 9,000 lines of~C++, of which about 1,000~lines are
shared with~Matt.
It~uses no external tools or libraries, and
its initial development took a year and a half.
It uses multidimensional dynamic programming to exactly compute the alignments
for which MRFy relies on stochastic search.
As a result, in the presence of complex beta-strand topologies, SMURF is
computationally intractable.
\item
MRFy is used to detect homologous proteins in the presence of paired
beta strands; it effectively supplants SMURF.
It~comprises about 1,200 lines of~Haskell.
\remark{HA HA HA.  It's over 3,500 lines now. Maybe around 800 lines
are tests or quickcheck properties?}
It~uses several external tools and libraries, of which the most
notable are Parsec, the BioHaskell bioinformatics library, and
the libraries \path{Data.MemoCombinators}, \path{Control.Parallel},
and \path{Data.Vector}.
Its initial development took about three months.
%%%% \remark{Others were mostly full-time work, but no need to distinguish.}
\end{itemize}
Like much research software, all three of these tools were written
 in~haste.
We~have experience modifying the older tools.

@@

Ideally we would use QuickCheck to find bugs, but as we mention in
\secref{awkward-quickcheck}, we found obstacles.
We~have now overcome these obstacles, but
we~sorely regret not doing so earlier.
In~light of our experience, we will institute a new programming practice:
whenever we introduce a new data type,
we~will write the instance of
\texttt{Arbitrary} right away, while
relevant invariants are still fresh in memory.
When invariants are not enforced by Haskell's static type system,
we~will write them as Haskell predicates.
We can then immediately run QuickCheck on each predicate, to verify
that our Arbitrary instance agrees with the predicate.
For each predicate~\texttt{p} we can also check
\mono{fmap (all p . shrink) arbitrary}.

@@

While the Haskell community offers many
enticing tools, libraries, and packages,
be aware that
not all of them are worth using.
Some are not ready for prime time, and some were once great but are
no~longer maintained.
The great packages, like \texttt{Data.MemoCombinators} and Parallel
Strategies, are truly great.
But~for amateurs, it's not always easy to tell the great packages from the wannabes
and the has-beens.
And even some of the great packages could be better documented, with
more examples.
